---
title: "Module 4: Set-Up of Dependent Variable"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    extra_dependencies: ["float"]
    number_sections: TRUE
    toc: true
    toc_depth: 1
#  html_document:
#    number_sections: TRUE
#    toc: true
#    toc_depth: 1
#    theme: flatly
#    highlight: zenburn
#  word_document:
#    number_sections: TRUE
#    toc: true
#    toc_depth: 1
urlcolor: magenta
geometry: margin=1in
header-includes:
   \usepackage[dvipsnames]{xcolor} 
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, fig.pos = 'H')
```

```{r, echo = FALSE, warning = FALSE}
path_pns_input_data <- Sys.getenv("path_pns_input_data")
path_pns_output_data <- Sys.getenv("path_pns_output_data")
path_pns_figures <- Sys.getenv("path_pns_figures")
```


\vspace{0.75cm}

**MODULE 4 GOAL:** By the end of this module, you will be able to:

* Learn the logic of constructing the Dependent Variable using scientific considerations (i.e., our motivating scientific question) and practical considerations (i.e., the data collection design).


#  Scientific Question

For convenience, let's display the scientific question we introduced in an earlier module.

**SCIENTIFIC QUESTION:** On average, is self-efficacy *at the current time point* associated with *the proximal occurrence of cigarette smoking* during the post-quit period?


# Implementation

**Reference file with R code:** `module-04.R`

We will zoom in on specific sections within the implementation in `module-04.R`.

An outer loop goes through each participant. For each participant, the inner loop goes through each of their rows in `dat_analysis`. Recall that, at this point, only Random EMAs having *any* response (i.e., `with_any_response=1`) are included in `dat_analysis`. Hence, for a particular participant, `total_random_ema` will be the total number of Random EMAs for which the participant provided *any* response.

```{r, eval = FALSE}
for(i in 1:total_participant_ids){
  
  # More code here
  
  for(j in 1:total_random_ema){
    
    # More code here
    
  }
}
```

\fboxrule.1em\fboxsep1em
\fcolorbox{black}{blue!3}{\color{black}
\begin{minipage}[c][2in][t]{6.3in}
\sffamily Draw Figure Here.
\end{minipage}}


The inner loop calculates `count_within_bounds`, our Dependent Variable.

```{r, eval = FALSE}
  for(j in 1:total_random_ema){
    current_lower_bound <- all_lower_bound[j]
    current_upper_bound <- all_upper_bound[j]
    
    # How many EMAs were launched between the two Random EMAs we
    # are looking at now?
    dat_within_bounds <- current_dat_smoking %>% 
      # Note the use of '>' instead of '>=' when checking against left end point
      # We do not include the number of reported cigarettes smoked in the left end point 
      # However, we will include the number of reported cigarettes smoked in the right end point
      filter((time_unixts > current_lower_bound) & (time_unixts <= current_upper_bound))
    
    number_within_bounds <- nrow(dat_within_bounds)
    # Only proceed with further calculations if we have at least one EMA
    if(number_within_bounds > 0){
      number_missing <- sum(is.na(dat_within_bounds$smoking_qty))
      # Only proceed with further calculations if there is no missing value in smoking_qty
      if(number_missing == 0){
        current_count_within_bounds <- sum(dat_within_bounds$smoking_qty)
      } # Mark end of IF STATEMENT
    } # Mark end of IF STATEMENT
    
    current_dat_analysis$count_within_bounds[j] <- current_count_within_bounds
  } # Mark end of FOR LOOP over Random EMAs having with_any_response=1
```

**What are the bounds?**

\fboxrule.1em\fboxsep1em
\fcolorbox{black}{blue!3}{\color{black}
\begin{minipage}[c][2in][t]{6.3in}
\sffamily Draw Figure Here.
\end{minipage}}


**Which EMAs within these bounds should we *not* use?**

\fboxrule.1em\fboxsep1em
\fcolorbox{black}{blue!3}{\color{black}
\begin{minipage}[c][2in][t]{6.3in}
\sffamily Draw Figure Here.
\end{minipage}}



**When may missing values come about in our Dependent Variable `count_within_bounds`?**

\fboxrule.1em\fboxsep1em
\fcolorbox{black}{blue!3}{\color{black}
\begin{minipage}[c][2in][t]{6.3in}
\sffamily Draw Figure Here.
\end{minipage}}



\fboxrule.1em\fboxsep1em
\fcolorbox{black}{red!5}{\color{black}
\begin{minipage}[0.30in]{6.3in}
\begin{center}BREAK: Any questions?\end{center}
\end{minipage}}




